---
title: "Child Malnutrition in Africa, with a Focus on Ghana"
subtitle: "From the continent, to Ghana, to its 16 regions"
author: "Kingsley Amegah"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5.5, fig.align = "center")
library(tidyverse); library(sf); library(scales); library(ggrepel)
theme_set(theme_minimal(base_size = 12))
ACC <- "#1b9e77"; HI <- "#d73027"
```

# Introduction

Malnutrition remains one of the defining health challenges in Africa, but the
picture is far from uniform. This report zooms in across three levels: where
**Ghana sits among African countries**, how malnutrition has **changed within
Ghana**, and how it is **distributed across Ghana's 16 regions** and social
groups. Ghana is an instructive case: a relative success story on chronic
undernutrition that still carries sharp internal inequalities.

The headline indicator is **stunting** (low height-for-age, the standard measure
of chronic child undernutrition), complemented by underweight, wasting,
overweight and anaemia.

# Data and methods

All data come from open sources, with no API keys required:

- **DHS Program API** - Ghana stunting by region and by wealth, education and
  residence, across survey rounds (1993-2022).
- **World Bank Open Data API** - malnutrition indicators for African countries
  and Ghana's national trend.
- **geoBoundaries** - Ghana's administrative (region) boundaries for mapping.

Analysis is descriptive and equity-focused: continental benchmarking, national
trends, a regional choropleth, and disparity ratios across social groups.

```{r load}
afr   <- read_csv("data/africa_stunting_latest.csv", show_col_types = FALSE)
trend <- read_csv("data/ghana_trend.csv", show_col_types = FALSE)
region<- read_csv("data/ghana_stunting_region.csv", show_col_types = FALSE)
equity<- read_csv("data/ghana_equity.csv", show_col_types = FALSE)
gh_sf <- st_read("data/ghana_adm1.geojson", quiet = TRUE) |>
  mutate(region = str_remove(shapeName, " Region$"))
```

# Africa context: where does Ghana stand?

```{r africa, fig.height=9}
afr <- afr |> mutate(is_ghana = iso3 == "GHA")
ghana_val <- afr$value[afr$iso3 == "GHA"]
afr_median <- median(afr$value)
ggplot(afr, aes(value, fct_reorder(country, value), fill = is_ghana)) +
  geom_col() +
  geom_vline(xintercept = afr_median, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c(`FALSE` = "grey75", `TRUE` = HI), guide = "none") +
  labs(title = "Child stunting across African countries (latest available)",
       subtitle = sprintf("Ghana = %.0f%% (red); African median = %.0f%% (dashed). Lower is better.",
                          ghana_val, afr_median),
       x = "Stunting prevalence (% of under-5s)", y = NULL)
```

Ghana's stunting prevalence (`r round(ghana_val)`%) is **well below the African
median** of `r round(afr_median)`%, placing it among the continent's stronger
performers on chronic undernutrition.

# Ghana over time: progress and the double burden

```{r trend}
trend |>
  filter(indicator %in% c("stunting", "underweight", "overweight", "severe_wasting")) |>
  ggplot(aes(year, value, colour = indicator)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Ghana: child malnutrition indicators over time",
       x = NULL, y = "% of children under 5", colour = NULL)
```

Chronic undernutrition (stunting, underweight) has fallen substantially, yet
**wasting has not disappeared and overweight persists** - the early signs of a
"double burden" of malnutrition that increasingly characterises middle-income
transitions.

# Within Ghana: the regional divide

We map stunting across Ghana's 16 regions using the most recent survey (2022).
DHS reports a mix of old and new regions; we select the 16 current regions and
match them to the boundary file.

```{r map}
# Map DHS 2022 leaf labels -> boundary region names
m <- c("..Western (post 2022)"="Western", "..Western North"="Western North",
       "Central"="Central", "Greater Accra"="Greater Accra",
       "..Volta (post 2022)"="Volta", "..Oti"="Oti", "Eastern"="Eastern",
       "Ashanti"="Ashanti", "..Ahafo"="Ahafo", "..Bono"="Bono",
       "..Bono East"="Bono East", "....Northern(post 2022)"="Northern",
       "....Savannah"="Savannah", "....Northeast"="North East",
       "..Upper East"="Upper East", "..Upper West"="Upper West")
reg22 <- region |> filter(survey_year == 2022, label %in% names(m)) |>
  mutate(region = m[label]) |> select(region, stunting = value)

map_df <- gh_sf |> left_join(reg22, by = "region")
ggplot(map_df) +
  geom_sf(aes(fill = stunting), colour = "white", linewidth = 0.3) +
  scale_fill_distiller(palette = "RdYlGn", direction = -1, name = "Stunting %") +
  geom_sf_text(aes(label = paste0(region, "\n", round(stunting), "%")), size = 2.4) +
  labs(title = "Child stunting by region, Ghana 2022",
       subtitle = "The north carries a far heavier burden than the south") +
  theme_void(base_size = 12)
```

```{r region-bar, fig.height=6}
reg22 |>
  ggplot(aes(stunting, fct_reorder(region, stunting))) +
  geom_col(fill = ACC) +
  geom_text(aes(label = paste0(round(stunting), "%")), hjust = -0.15, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Stunting by region, Ghana 2022 (ranked)",
       x = "Stunting (%)", y = NULL)
```

```{r region-stats, echo=FALSE}
hi <- reg22 |> slice_max(stunting, n = 1)
lo <- reg22 |> slice_min(stunting, n = 1)
```

The gap is stark: stunting ranges from about **`r round(lo$stunting)`% in
`r lo$region`** to **`r round(hi$stunting)`% in `r hi$region`** - a
`r round(hi$stunting/lo$stunting, 1)`-fold difference between the best and
worst regions, reflecting Ghana's well-documented north-south divide.

# Inequality by wealth, education and residence

```{r equity}
eq <- equity |>
  filter(survey_year == 2022, indicator == "stunting",
         category %in% c("Wealth quintile", "Education", "Residence", "Sex")) |>
  mutate(label = factor(label, levels = c(
    "Lowest","Second","Middle","Fourth","Highest",
    "No education","Primary","Secondary","Higher",
    "Rural","Urban","Male","Female")))
ggplot(eq, aes(label, value, fill = category)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(value), "%")), vjust = -0.3, size = 3) +
  facet_wrap(~ category, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Stunting by social group, Ghana 2022",
       x = NULL, y = "Stunting (%)")
```

```{r equity-stats, echo=FALSE}
gv <- function(cat, lab) eq$value[eq$category==cat & eq$label==lab]
w_ratio <- gv("Wealth quintile","Lowest")/gv("Wealth quintile","Highest")
e_ratio <- gv("Education","No education")/gv("Education","Higher")
```

The social gradient is steep. Children in the **poorest** wealth quintile are
about **`r round(w_ratio,1)` times** as likely to be stunted as those in the
**richest** (`r round(gv("Wealth quintile","Lowest"))`% vs
`r round(gv("Wealth quintile","Highest"))`%), and children of mothers with
**no education** are around **`r round(e_ratio,1)` times** as likely as those
whose mothers have higher education. Rural children fare worse than urban ones.
These gaps show that national averages hide who is being left behind.

# Conclusion and limitations

**Conclusion.** Ghana is a relative success on chronic child undernutrition,
outperforming most African countries and reducing stunting over two decades. But
the national average conceals a sharp **north-south divide** and steep
**wealth and education gradients**. Progress now depends less on national growth
than on **reaching the poorest regions and households** - and emerging overweight
signals a coming double burden.

**Limitations.** Survey estimates carry sampling error and are collected only
every few years; region definitions changed in 2019, complicating long-run
regional comparisons; and this is an ecological, descriptive analysis, so
associations should not be read as causal.

---

*Built in R (tidyverse, sf, ggplot2) from DHS, World Bank and geoBoundaries open
data. Author: Kingsley Amegah.*
